Parametricity Consequences of GADTs Representations

### 2021-02-23 20-gadts/GADT-story.tex
work on gadt-story

\section{Introduction}
{\color{blue} map vs. fmap throughout}

@@

By contrast with the ADT $\mathsf{List\, a}$, where the type parameter
$\mathsf{a}$ is integral to the type being defined, the type parameter
$\mathsf{a}$ appears in both $\mathsf{PTree\,a}$ and $\mathsf{Seq\,a}$
as a ``dummy'' parameter used only to give the kind \verb|* -> *| of
the type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
explicitly captured in the alternative Haskell syntax {\color{blue}
    Correct?}
\[\begin{array}{l}
\mathsf{data\, Seq :: * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}\]
A GADT --- even a nested type --- thus does not define a {\em family
  of inductive types}, one for each type argument, like an ADT does,
but instead defines an entire family of types that must be constructed
simultaneously. That is, a GADT defines an {\em inductive family of
  types}.

@@

{\color{red} HERE!}

The syntax of GADTs is consonant with programmers' frequent use of
them in situations where type uniformity is specifically to be
eschewed.  {\color{blue} Typical applications of GADTs include generic
  programming, modeling programming languages via higher-order
  abstract syntax, maintaining invariants in data structures, and
  expressing constraints in embedded domain-specific languages. GADTs
  have also been used to implement tagless
  interpreters~\cite{pl04,pr06,pvww06} by trading the definition of a
  universal value domain for a direct specification of the property of
  being a value. Other applications are described in,
  e.g.,~\cite{js15,rou06,MORE,UPDATED}. OTHER SITUATIONS NEEDING
  TYPE-BASED DISPATCH.}

GADTs are useful in such situations precisely because generalizing the
type arguments in the instances of a GADT appearing as the return
types of its data constructors entails a strong connection between a
GADTs shape and the data it contains.  With ADTs, we first choose the
shape of the container and then fill that container with data of
whatever type we like. Critically, the choice of shape is independent
of the data to be stored. With GADTs, however, the shape of the
container may actually {\em depend} on (the type of) the data to be
contained. On other words, the shape and the data are not independent!
For example, while $\mathsf{Const}$ can create data of any shape
$\mathsf{Seq\,a}$, $\mathsf{Pair}$ can produce data of shape
$\mathsf{Seq\,a}$ where $\mathsf{a}$ is a pair type. As a result, it
may not always be possible to modify a GADT's data while retaining its
shape because changing (the type of) the data may actually change the
shape.

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, a
GADT calculus must support an equality type $\mathsf{Equal}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
function $\mathit{Equal}$. {\color{blue} This is the least commitment
  thing Church encodings require.} Such a type can either be taken as
primitive, as in
\[\begin{array}{l}
\mathsf{data\, Equal :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,a\,a}\\
\end{array}\]
\noindent
or
represented by its own Church encoding
\[\mathsf{Equal\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Equal}$, the GADT
$\mathsf{Seq}$ can then be represented as its Church encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
  \to (\forall a\,b\,c.\,Equal\,a\,(b\times c) \to f\,b \to f\,c \to
  f\,a) \to f\,a}\]
in a higher-kinded calculus such as $F_\omega$.

Importantly, $\mathit{Equal}\,a$ cannot be made into a functor, or,
equivalently, $\mathsf{Equal\,a}$ cannot be made an instance of
Haskell's $\mathsf{Functor}$ class: if $\mathsf{Equal\,a}$ supported a
function
\[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
  Equal\,a\,c}\]
then defining
\[\begin{array}{l}
\mathsf{eqElim :: Equal\, a\, b \to b \to a}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct the element 
\[\begin{array}{l}
\mathsf{emptyFromMapEquala :: (\forall b\, c.\, (b \to c) \to Equal\, a\,
  b \to Equal\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEquala \,map_{Equal\,a} = eqElim\, (map_{Equal\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
of type $\mathsf{Void}$, which is impossible. It follows immediately
from this contradiction that $\mathsf{Seq}$ cannot support a map
function either. This is already intuitively evident from the Haskell
definition of $\mathsf{Seq}$, though. ALREADY USED??? Indeed, the clause of
$\mathsf{map}$ for the $\mathsf{Pair}$ constructor should feed
$\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and a
term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$ and
$\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$ for
some appropriately typed terms $\mathsf{t_1'}$ and
$\mathsf{t_2'}$. But it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
then we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions.

@@

The primitive representations of {\color{blue} polynomial} GADTs
capture in syntax the fact that GADTs can be considered as fixpoints
of higher-order functors. Syntactically, the use of higher-order
funct{\em ions} is essential, since the type arguments to the GADT
being defined are not uniform across all of its instances in the types
of its data constructors. Semantically, the use of higher-order
funct{\em ors} is essential, as for ADTs, to the very existence of the
fixpoints being computed~\cite{tfca}.
%We will consider GADTs' semantics in more detail in
%Section~\ref{sec:GADT-semantics}.

Consider, again, the GADT $\mathsf{Seq}$ from the start of
Section~\ref{sec:gadts}. Because its type argument varies in the
instances of $\mathsf{Seq}$ appearing in the types of its data
constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
any first-order fixpoint equation. It can, however, be viewed as a
solution to the higher-order fixpoint equation
\[H\,f\,a = a \,+\,Lan_{\lambda a b. a \times b}\, \lambda
a b. f a \times f b\] where $Lan_K \,G$ is the left Kan extension of
$G$ along $K$.  The left Kan extension $Lan_K \,G : {\cal E} \to
{\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along the functor
$K : {\cal C} \to {\cal E}$ is the best functorial approximation to
$G$ that factors through $K$. Here, ``best functorial approximation''
means that $Lan_K \,G$ is the smallest functor that both extends the
image of $K$ to $\cal D$ and coincides with $G$ on $\cal C$, in the
sense that, for any other such functor $H$, there is a morphism of
functors (i.e., a natural transformation) from $Lan_K \,G$ to
$H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $G : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $G$ along $K$} is a
functor $\mathit{Lan}_K\,G : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : G \to (\mathit{Lan}_K\,G) \circ K$
such that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : G \to H \circ K$, there exists a unique
natural transformation $\delta : \mathit{Lan}_K\,G \to H$ such that $(\mu
K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{G}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,G}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
\end{tikzcd}\]
\end{definition}

@@

@@

>>>>

\section{Introduction}\label{sec:intro}
{\color{blue} map vs. fmap throughout.} {\color{blue} Add polynomial
  throughout?}

@@

By contrast with the ADT $\mathsf{List\, a}$, where the type parameter
$\mathsf{a}$ is integral to the type being defined, the type parameter
$\mathsf{a}$ appears in both $\mathsf{PTree\,a}$ and $\mathsf{Seq\,a}$
as a ``dummy'' parameter used only to give the kind \verb|* -> *| of
the type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
explicitly captured in the alternative ``kind signature'' Haskell
syntax
\[\begin{array}{l}
\mathsf{data\, Seq :: * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}\]
A GADT --- even a nested type --- thus does not define a {\em family
  of inductive types}, one for each type argument, like an ADT does,
but instead defines an entire family of types that must be constructed
simultaneously. That is, a GADT defines an {\em inductive family of
  types}.

@@

The syntax of GADTs is consistent with programmers' use of them in
situations where different behaviors at different type instances is
desired. This is achieved by allowing the progammer to give the type
signatures of the different data constructors independently, and then
using pattern matching to force the desired type refinement.  Typical
applications of GADTs include generic programming, modeling
programming languages via higher-order abstract syntax, maintaining
invariants in data structures, and expressing constraints in embedded
domain-specific languages. They have also been used, e.g., to
implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
memory performance~\cite{min15}, and to design APIs~\cite{pen20}.

Allowing non-variable type arguments in the return types of a GADT's
data constructors establishes a strong connection between a GADT's
shape and the data it contains. With ADTs, we first choose the shape
of the container and then fill that container with data of whatever
type we like. Critically, the choice of shape is independent of the
data to be stored. With GADTs, however, the shape of the container may
actually {\em depend} on (the type of) the data to be contained. For
example, $\mathsf{Const}$ can create data of any shape
$\mathsf{Seq\,a}$, but $\mathsf{Pair}$ can produce data of shape
$\mathsf{Seq\,a}$ only if $\mathsf{a}$ is a pair type. As a result,
modifying the data in a GADT may actually change the shape of --- or
may even destroy! --- the GADT.

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, it is
essential that a GADT calculus support an equality type
$\mathsf{Equal}$ that is inhabited when its two type arguments are the
same and is uninhabited otherwise. That is, the type $\mathsf{Equal}$
must be the syntactic reflection of semantic equality function
$\mathit{Equal}$. The type $\mathsf{Equal}$ can either be defined via
GADT syntax, as in
\[\begin{array}{l}
\mathsf{data\, Equal\,a\,b\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,c\,c}\\
\end{array}\]
\noindent
or, equivalently, represented by its own Church encoding
\[\mathsf{Equal\,a\,b} = \mathsf{\forall f.\, (\forall c.\,
  f\,c\,c) \to f\,a\,b}\]
The GADT $\mathsf{Seq}$ can then be represented as its Church encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
  \to (\forall a\,b\,c.\,Equal\,a\,(b\times c) \to f\,b \to f\,c \to
  f\,a) \to f\,a}\]
in a higher-kinded calculus such as $F_\omega$.

Importantly, the function $\mathit{Equal}\,a$ cannot be made into a
functor. Equivalently, $\mathsf{Equal\,a}$ (in either representation)
cannot be made an instance of Haskell's $\mathsf{Functor}$
class. Indeed, if $\mathsf{Equal\,a}$ supported a function
\[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
  Equal\,a\,c}\]
then defining
\[\begin{array}{l}
\mathsf{eqElim :: Equal\, a\, b \to b \to a}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct an element 
\[\begin{array}{l}
\mathsf{emptyFromMapEquala :: (\forall b\, c.\, (b \to c) \to Equal\, a\,
  b \to Equal\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEquala \,map_{Equal\,a} = eqElim\, (map_{Equal\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
of type $\mathsf{Void}$. {\color{blue} Use same type conventions as in
  Intro.} But this should be impossible. It is immediate that no GADT
that is not actually a nested type can support a $\mathsf{map}$
function either. In the case of $\mathsf{Seq}$, this is already
intuitively evident from its Haskell definition as discussed in
Section~\ref{sec:intro}.

@@

{\color{red} HERE!!!}

Primitive representations of GADTs capture in syntax the fact that
GADTs can, like ADTs, be considered as fixpoints of higher-order
functors. Syntactically, the use of higher-order funct{\em ions} is
essential, since the type arguments to the GADT being defined are not
uniform across all of its instances in the types of its data
constructors. Semantically, the use of higher-order funct{\em ors} is
essential, as for ADTs, to the very existence of the fixpoints being
computed~\cite{tfca}.
%We will consider GADTs' semantics in more detail in
%Section~\ref{sec:GADT-semantics}.

Consider, again, the GADT $\mathsf{Seq}$. Because its type argument
varies in the instances of $\mathsf{Seq}$ appearing in the types of
its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ simply cannot be
characterized by any first-order fixpoint equation.  {\color{blue}
  Compare with ADT situation.}  It can, however, be viewed as a
solution to the higher-order fixpoint equation
\[H\,f\,a = a \,+\,Lan_{\lambda a b. a \times b}\, \lambda
a b. f a \times f b\] where $Lan_K \,G$ is the left Kan extension of
$G$ along $K$.  In general, the left Kan extension $Lan_K \,G : {\cal
  E} \to {\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along
the functor $K : {\cal C} \to {\cal E}$ is the best functorial
approximation to $G$ that factors through $K$. Here, ``best functorial
approximation'' means that $Lan_K \,G$ is the smallest functor that
both extends the image of $K$ to $\cal D$ and coincides with $G$ on
$\cal C$, in the sense that, for any other such functor $H$, there is
a morphism of functors (i.e., a natural transformation) from $Lan_K
\,G$ to $H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $G : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $G$ along $K$} is a
functor $\mathit{Lan}_K\,G : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : G \to (\mathit{Lan}_K\,G) \circ K$
such that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : G \to H \circ K$, there exists a unique
natural transformation $\delta : \mathit{Lan}_K\,G \to H$ such that $(\mu
K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{G}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,G}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
\end{tikzcd}\]
\end{definition}

@@

\subsection{Comparison with Stephanie and Dmitrios}

@@

GADTs don't really fit into his scheme because the two reps of GADTs
aren't the same. Bob's stuff only works for functorial data types of
any kind -- so not GADTs. So last section is misleading.
